Homogenization of periodic and random ferroelectric-dielectric composites

### 2019-03-20 paper/main.tex
dev

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
  We investigate the homogenized permittivity of ferroelectric-dielectric mixtures under
  a static electric field. A refined model is used to take into account the coupling
  between the electrostatic problem and the electric field dependent permittivity of the
  ferroelectric material. Periodic and random structures in two dimensions are investigated and
  we study the effective permittivity, losses, electrically induced anisotropy and tunability
   of the homogenized materials.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{fmtext}
  \section{Introduction}

  Ferroelectric materials play a crucial role in tunable
  microwave devices, with typical applications including antenna beam steering,
  phase shifters, tunable power splitters, filters, voltage controlled oscillators and
  matching networks \cite{Tagantsev2018}. Both bulk ceramics and thin films have
  been employed to design frequency agile components \cite{vendik1999ferroelectric, lancaster1998thin,
  xi2000oxide} and metamaterials \cite{Hand2008, Zhao2008}. The main reason of using
  ferroelectric materials is their strong dependence of their permittivity $\varepsilon$
  on an applied electric field $E$, which is measured by their tunability defined as $n = \varepsilon(0)/\varepsilon(E)$.
  The key requirements for antenna and microwave applications are large tunability and low losses.
  These two characteristics are correlated and one has to find a trade-off for optimal device
  performance, which can be quantified by the so called commutation quality factor
  $K = (n -1)^2/(n\, \tan\delta(0)\,\tan\delta(E))$, where $\tan\delta$ is the loss tangent.\\
  These materials have usually high permittivity values,
  often leading to slow response time and impedance mismatch, which can be an issue in some practical
  applications. Thus it has been considered to mix ferroelectric ceramics to low-index and
  low-loss non-tunable dielectrics in order to reduce both permittivity
\end{fmtext}

@@

\maketitle
\noindent values and losses.
The effective parameters of those composites have been investigated \cite{Sherman2006,
jylha2008tunability, Sherman2004, ASTAFIEV20032381} and it has been found that the
permittivity can be greatly reduced while losses and tunability are much less sensitive to the
dielectric phase addition.\\
This study investigates the effective permittivity of dielectric/ferroelectric composites
by using a two-scale convergence method \cite{Allaire92, Guenneau2000}.
The originality lies in the fact that a fully coupling model is employed to
calculate the electrostatic field distribution when a uniform biasing field is
applied on the structures, which will result in a local modification of the permittivity
in the ferroelectric phase due to the microstructure. As compared to a simple uncoupled model where the
ferroelectric phase is only modified through the biasing field,
the resulting effective permittivity, dielectric losses, tunability and
anisotropy significantly differ. \\
We study first a two-dimensional model of metamaterials made of parallel rods with circular cross section,
considering both square periodic arrays and random distributions within the unit cell of
a periodic array.

@@

We use barium strontium titanate (BST) as our ferroelectric material. Measurements
have been carried out at electrostatics and at microwave frequencies, and the normalized permittivity value
as a function of biasing field are reported on \fig{fig1}.\\
To describe the permittivity, we make use of the Landau potential
given by $F(P,E) = F_0 +  a P^2/2 + b P^4/4 + cP^6/6 - EP$, where $E$ is
the applied electric field and $P$ is the polarization \cite{landau_electrodynamics_2013, Zhou2008}. Variations of the
permittivity with the temperature can be taken into account through the
 coefficients $a$, $b$ and $c$, but we assume we are working at a constant
room temperature. We further assume that the material is not subject to any stress, so that the variation
of permittivity due to mechanical constraints is irrelevant.
The equation of state $$\frac{\partial F (P, E)}{\partial P}   = a P_0 + b P_0^3 + c P_0^5 - E = 0$$ gives the
dependence of the polarization on the applied electric field,
with $P_0$ being the equilibrium polarization.
Along the direction of a uniform applied electric field, the relative permittivity is given by:
\begin{equation}
  \epsf(E) = \left[\frac{\partial^2 F (P, E)}{\partial P^2} \right]^{-1} = \frac{\epsf(0)}{1 + \alpha P_0^2 + \beta P_0^4},
  \label{eq_epsf}
\end{equation}
where $\epsf(0)=1/a$, $\alpha=3b/a$ and $\beta=5c/a$. The fitting parameters are given in Table \ref{table_params_fit}.
As the norm of the field increases, the permittivity decreases with a characteristic
bell curve typical for a ferroelectric material in its paraelectric state.
\begin{figure}[!t]
\centering
\includegraphics[width=0.8\columnwidth]{epsilon_fit}
\caption{Variation of the ferroelectric permittivity as a function of the
 applied electric field (dots: measurements, dashed lines: fit to
 formula~(\ref{eq_epsf})), for the static case (green) and at microwave frequencies
  (orange, $f=3.8$ GHz). The fitting parameters are given in Table \ref{table_params_fit}.}
\label{fig1}
\end{figure}
Furthermore, assuming the crystalline principal axes of the ferroelectric material
are oriented in the coordinate directions, and that the diagonal components of the permittivity
tensor are only function of the corresponding bias electric field components \cite{Krowne2002}, we have:
\begin{equation}
  \epsftens (\B E) =
\left(
\begin{matrix}
\epsf_{xx}(E_x) & 0 & 0 \cr
0 & \epsf_{yy}(E_y) & 0 \cr
0 & 0 & \epsf_{zz}(E_z) \cr

@@

% -------------------------------------------------------------------------------
\subsection{Homogenization}
The effective permittivity for TM polarization is calculated
using two scale homogenization technique \cite{Allaire92, Guenneau2000}.
For this purpose, one has to find the solutions
$\psi_j$ of two annex problems $\mathcal P_j$, $j=\{1, 2\}$:
\begin{equation}
\div \left[ \xitens \grad(\psi_j + r_j) \right] = 0,
\label{eq_hom_annex}
\end{equation}
where $\B r = (x, y)^T$ is the position vector in the $xy$ plane and
$\xitens=\epstens^{\rm T}/ {\rm det }(\epstens)$.
The homogenized coefficient $\xihom$ is obtained with:
\begin{equation}
\xihom = \langle \xitens \rangle + \B \phi,
\label{eq_hom}
\end{equation}
where $\langle . \rangle$ denotes the mean value over the unit cell.
The elements of the matrix $\B \phi$ represent correction terms and
are given by $\B \phi_{ij} = \langle \xitens \grad \psi_i \rangle_j$.
Finally the homogenized permittivity  can be calculated using $\epshom=\xihom^{\rm T}/ {\rm det }(\xihom)$.\\
Note that the TE case, which we shall not study here as no coupling happens, is trivial since
the homogenized permittivity is simply the average of the permittivity in the unit cell:
$\epshom = \langle \epstens \rangle$.

% -------------------------------------------------------------------------------
\subsection{Numerical setup}
Equations~(\ref{eq_elstat}) and (\ref{eq_hom_annex}) are solved with a Finite Element
Method using the open source packages Gmsh \cite{gmsh} and GetDP \cite{getdp}.
In both cases we use a square unit cell $\Omega$ of length $d$ with periodic boundary
conditions along $x$ and $y$. Second order Lagrange elements are used and the
solution is computed with a direct solver (MUMPS \cite{MUMPS}).
% The source code and
% data necessary to reproduce the results reported in this paper are freely available
% online.

@@

We have studied the homogenized properties of dielectric/ferroelectric mixtures
using a rigorous model that take into account the coupling between the electrostatic
field distribution and the field dependant ferroelectric permittivity tensor. After
convergence of the coupled problem, the effective permittivity tensor is calculated using
two scale convergence homogenization theory.
The results obtained by this model differ significantly from a simple assumption that
the permittivity of the ferroelectric respond just to the uniform biasing field.
We have considered both periodic and random arrays
of dielectric rods in a ferroelectric matrix in 2D, and studied their effective properties
for TM polarization as a function of dielectric concentration and bias field.
Importantly, adding more low index and low loss dielectric allows to decrease
the overall permittivity significantly and slightly lower the losses. On the other hand,
the higher the dielectric concentration the smaller the tunability would be while
anisotropy gets higher as the concentration decreases.
The properties of the composites are affected by multiple factors:
 geometry and the spatially dependent electric field that will induce locally a tunable, anisotropic
 response in the ferroelectric phase depending on its amplitude and direction.
 This suggest that the performances of the composites
 may be enhanced by distributing the two phases in an optimal way to get high
  tunability and low losses. Further work in that direction is needed as well as
  extending this study to 3D media.
 Finally, because the permittivity of the dielectric is much smaller than the ferroelectric one,
 it would be of great interest to use high contrast homogenization theory
 \cite{BOUCHITTE2004377, cherednichenko_cooper_2015} to
 study this kind of mixtures.
 This would reveal the frequency dependant artificial magnetism due to "micro-resonances"
 in the high index phase and potentially lead to composites with tunable effective permeability.

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
  We investigate the homogenized permittivity of ferroelectric-dielectric mixtures under
  a static electric field. A refined model is used to take into account the coupling
  between the electrostatic problem and the electric field dependent permittivity of the
  ferroelectric material. Periodic and random structures in two dimensions are investigated and
  we study the effective permittivity, losses, electrically induced anisotropy and tunability
   of those metamaterials.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{fmtext}
  \section{Introduction}
  Ferroelectric materials play a crucial role in reconfigurable 
  microwave devices, with typical applications including antenna beam steering,
  phase shifters, tunable power splitters, filters, voltage controlled oscillators and
  matching networks \cite{tagantsev_ferroelectric_2018}. Both bulk ceramics and thin films have
  been employed to design frequency agile components \cite{vendik_ferroelectric_1999, lancaster_thin-film_1998,xi_oxide_2000} and metamaterials \cite{hand_frequency_2008, zhao_experimental_2008}. The main reason of using
  ferroelectric materials is their strong dependence of their permittivity $\varepsilon$
  on an applied electric field $E$, which is measured by their tunability defined as $n = \varepsilon(0)/\varepsilon(E)$.
  The key requirements for antenna and microwave applications are large tunability and low losses.
  These two characteristics are correlated and one has to find a trade-off for optimal device
  performance, which can be quantified by the so called commutation quality factor
  $K = (n -1)^2/(n\, \tan\delta(0)\,\tan\delta(E))$, where $\tan\delta$ is the loss tangent.\\
  These materials have usually high permittivity values,
  often leading to slow response time and impedance mismatch, which can be an issue in some practical
  applications. Thus it has been considered to mix ferroelectric ceramics to low-index and
  low-loss non-tunable dielectrics in order to reduce both permittivity
\end{fmtext}

@@

\maketitle
\noindent values and losses.
The effective parameters of those composites have been investigated \cite{sherman_ferroelectric-dielectric_2006, jylha_tunability_2008, sherman_tunability_2004, astafiev_can_2003} and it has been found that the
permittivity can be greatly reduced while losses and tunability are much less sensitive to the
dielectric phase addition.\\
This study investigates the effective permittivity of dielectric/ferroelectric composites
by using a two-scale convergence method \cite{allaire_homogenization_1992, guenneau_homogenization_2000}.
The originality lies in the fact that a fully coupling model is employed to
calculate the electrostatic field distribution when a uniform biasing field is
applied on the structures, which will result in a local modification of the permittivity
in the ferroelectric phase due to the microstructure. As compared to a simple uncoupled model where the
ferroelectric phase is only modified through the biasing field,
the resulting effective permittivity, dielectric losses, tunability and
anisotropy significantly differ. \\
We study first a two-dimensional model of metamaterials made of parallel rods with circular cross section,
considering both square periodic arrays and random distributions within the unit cell of
a periodic array.

@@

We use barium strontium titanate (BST) as our ferroelectric material. Measurements
have been carried out at electrostatics and at microwave frequencies, and the normalized permittivity value
as a function of biasing field are reported on \fig{fig1}.\\
To describe the permittivity, we make use of the Landau potential
given by $F(P,E) = F_0 +  a P^2/2 + b P^4/4 + cP^6/6 - EP$, where $E$ is
the applied electric field and $P$ is the polarization \cite{landau_electrodynamics_2013, zhou_dielectric_2008}. Variations of the
permittivity with the temperature can be taken into account through the
 coefficients $a$, $b$ and $c$, but we assume we are working at a constant
room temperature. We further assume that the material is not subject to any stress, so that the variation
of permittivity due to mechanical constraints is irrelevant.
The equation of state $$\frac{\partial F (P, E)}{\partial P}   = a P_0 + b P_0^3 + c P_0^5 - E = 0$$ gives the
dependence of the polarization on the applied electric field,
with $P_0$ being the equilibrium polarization.
Along the direction of a uniform applied electric field, the relative permittivity is given by:
\begin{equation}
  \epsf(E) = \left[\frac{\partial^2 F (P, E)}{\partial P^2} \right]^{-1} = \frac{\epsf(0)}{1 + \alpha P_0^2 + \beta P_0^4},
  \label{eq_epsf}
\end{equation}
where $\epsf(0)=1/a$, $\alpha=3b/a$ and $\beta=5c/a$. The fitting parameters are given in Table \ref{table_params_fit}.
As the norm of the field increases, the permittivity decreases with a characteristic
bell curve typical for a ferroelectric material in its paraelectric state.
\begin{figure}[!t]
\centering
\includegraphics[width=0.8\columnwidth]{epsilon_fit}
\caption{Variation of the ferroelectric permittivity as a function of the
 applied electric field (dots: measurements, dashed lines: fit to
 formula~(\ref{eq_epsf})), for the static case (green) and at microwave frequencies
  (orange, $f=3.8$ GHz). The fitting parameters are given in Table \ref{table_params_fit}.}
\label{fig1}
\end{figure}
Furthermore, assuming the crystalline principal axes of the ferroelectric material
are oriented in the coordinate directions, and that the diagonal components of the permittivity
tensor are only function of the corresponding bias electric field components \cite{krowne_anisotropic_2002}, we have:
\begin{equation}
  \epsftens (\B E) =
\left(
\begin{matrix}
\epsf_{xx}(E_x) & 0 & 0 \cr
0 & \epsf_{yy}(E_y) & 0 \cr
0 & 0 & \epsf_{zz}(E_z) \cr

@@

% -------------------------------------------------------------------------------
\subsection{Homogenization}
The effective permittivity for TM polarization is calculated
using two scale homogenization technique \cite{allaire_homogenization_1992, guenneau_homogenization_2000}.
For this purpose, one has to find the solutions
$\psi_j$ of two annex problems $\mathcal P_j$, $j=\{1, 2\}$:
\begin{equation}
\div \left[ \xitens \grad(\psi_j + r_j) \right] = 0,
\label{eq_hom_annex}
\end{equation}
where $\B r = (x, y)^{\rm T}$ is the position vector in the $xy$ plane and
$\xitens=\epstens^{\rm T}/ {\rm det }(\epstens)$.
The homogenized coefficient $\xihom$ is obtained with:
\begin{equation}
\xihom = \langle \xitens \rangle + \B \phi,
\label{eq_hom}
\end{equation}
where $\langle . \rangle$ denotes the mean value over the unit cell.
The elements of the matrix $\B \phi$ represent correction terms and
are given by $\B \phi_{ij} = \langle \xitens \grad \psi_i \rangle_j$.
Finally the homogenized permittivity  can be calculated using $\epshom=\xihom^{\rm T}/ {\rm det }(\xihom)$.\\
Note that the TE case, which we shall not study here as no coupling happens, is trivial since
the homogenized permittivity is simply the average of the permittivity in the unit cell:
$\epshom = \langle \epstens \rangle$.

% -------------------------------------------------------------------------------
\subsection{Numerical setup}
Equations~(\ref{eq_elstat}) and (\ref{eq_hom_annex}) are solved with a Finite Element
Method using the open source packages Gmsh \cite{geuzaine_gmsh:_2009} and GetDP \cite{dular_general_1998}.
In both cases we use a square unit cell $\Omega$ of length $d$ with periodic boundary
conditions along $x$ and $y$. Second order Lagrange elements are used and the
solution is computed with a direct solver (MUMPS \cite{amestoy_fully_2001}).
% The source code and
% data necessary to reproduce the results reported in this paper are freely available
% online.

@@

We have studied the homogenized properties of dielectric/ferroelectric mixtures
using a rigorous model that take into account the coupling between the electrostatic
field distribution and the field dependant ferroelectric permittivity tensor. After
convergence of the coupled problem, the effective permittivity tensor is calculated using
two scale convergence homogenization theory.
The results obtained by this model differ significantly from a simple assumption that
the permittivity of the ferroelectric respond just to the uniform biasing field.
We have considered both periodic and random arrays
of dielectric rods in a ferroelectric matrix in 2D, and studied their effective properties
for TM polarization as a function of dielectric concentration and bias field.
Importantly, adding more low index and low loss dielectric allows to decrease
the overall permittivity significantly and slightly lower the losses. On the other hand,
the higher the dielectric concentration the smaller the tunability would be while
anisotropy gets higher as the concentration decreases.
The properties of the composites are affected by multiple factors:
 geometry and the spatially dependent electric field that will induce locally a tunable, anisotropic
 response in the ferroelectric phase depending on its amplitude and direction.
 This suggest that the performances of the composites
 may be enhanced by distributing the two phases in an optimal way to get high
  tunability and low losses. Further work in that direction is needed as well as
  extending this study to 3D media.
 Finally, because the permittivity of the dielectric is much smaller than the ferroelectric one,
 it would be of great interest to use high contrast homogenization theory
 \cite{bouchitte_homogenization_2004, cherednichenko_homogenization_2015} to
 study this kind of mixtures.
 This would reveal the frequency dependant artificial magnetism due to "micro-resonances"
 in the high index phase and potentially lead to composites with tunable effective permeability.
